---
title: "Summary"
author: "Yang Liu"
date: 
format: html
execute:
  echo: false
  warning: false
editor: visual
---

Title

## Background

Educational Intervention to Improve midwife students Skills in Serious Illness Communication

## Study population

midwife students n=17

```         
```

## Objective

compare their readiness and comfortability delivering serious news before and after training,

## Primary outcome

## Variables of interest

-7 skills

-comfortability

## Data management

```{r}
library(tidyverse)
library(rio)
library(readxl)
```

```{r}
# Read the data 
presurvey <- read_excel("rawdata/presurvey_num.xlsx", skip = 1)  
postsurvey <-read_excel("rawdata/postsurvey_num.xlsx", skip = 1)
#data structure
str(presurvey)
str(postsurvey)
# Replace empty cells with NA in presurvey dataset with base R
presurvey[] <- lapply(presurvey, function(x) {
  if (is.numeric(x) || is.integer(x)) {
    x[x == ""] <- NA
  } else if (is.character(x)) {
    x[x == ""] <- NA_character_
  } else if (is.factor(x)) {
    levels(x)[levels(x) == ""] <- NA
  }
  return(x)
})

# Replace empty cells with NA in postsurvey dataset with base R
postsurvey[] <- lapply(postsurvey, function(x) {
  if (is.numeric(x) || is.integer(x)) {
    x[x == ""] <- NA
  } else if (is.character(x)) {
    x[x == ""] <- NA_character_
  } else if (is.factor(x)) {
    levels(x)[levels(x) == ""] <- NA
  }
  return(x)
})
#check missing
check_na <- function(df) {
  # Calculate total number of NA values in the data frame
  total_na <- sum(is.na(df))
  
  # Calculate the number of NA values for each column
  na_per_column <- colSums(is.na(df))
  
  # Create a summary list
  na_summary <- list(
    total_na = total_na,
    na_per_column = na_per_column
  )
  
  return(na_summary)
}
check_na(presurvey)
check_na(postsurvey)
# RENAME varaible
presurvey <- presurvey %>%
  rename_with(
    ~ "id",                       
    starts_with("Please enter the last 4 digits"))
postsurvey <- postsurvey %>%
  rename_with(
    ~ "id",                       
    starts_with("Please enter the last 4 digits"))
# replace NA
presurvey$id[is.na(presurvey$id)] <- 2995
presurvey$id[presurvey$id == "0162"] <- "0612"

# Perform the matching
  complete_data <- merge(
    x = presurvey,
    y = postsurvey,
    by = "id",
    suffixes = c("_pre", "_post")
  )
#save complete dataset
export(complete_data, "complete_data.rds")
```

## Descriptive summary

```{r}
#| echo: false
library(tidyverse)
library(gtsummary)
library(rio)
# read rds
complete_data <- readRDS("complete_data.rds")
# descriptive summary
# add Pre: to pre-workshop questions
complete_data <- complete_data %>%
  rename(
    `Pre: Convey serious news about a patient's illness to the patient or family` = `Convey serious news about a patient's illness to the patient or family`,
    `Pre: Convey prognosis to a patient or family member` = `Convey prognosis to a patient or family member`,
    `Pre: Recognize emotion and respond with empathy to a patient or family member` = `Recognize emotion and respond with empathy to a patient or family member`,
    `Pre: Lead a family meeting` = `Lead a family meeting`,
    `Pre: Manage conflict that arises during a difficult discussion` = `Manage conflict that arises during a difficult discussion`,
    `Pre: Respond to patients who deny the seriousness of their illness` = `Respond to patients who deny the seriousness of their illness`,
    `Pre: Respond to patients who want treatments you believe are not indicated` = `Respond to patients who want treatments you believe are not indicated`,
    `Pre: Discuss religious or spiritual issues with a patient` = `Discuss religious or spiritual issues with a patient`,
    `Pre: Elicit a patient's personal values and preferences` = `Elicit a patient's personal values and preferences`
  )
# recode likert scale data 1-5 into binary 0,1
# Function to recode values using tidyverse
recode_values_tidy <- function(x) {
  if_else(x %in% 1:3, 0, if_else(x %in% 4:5, 1, NA_integer_))
}
# Recode the values for specified columns using dplyr and across
complete_data1 <- complete_data %>%
  mutate(across(
    contains("Pre:"),
    recode_values_tidy
  )) %>%
  mutate(across(
    contains("BEFORE TRAINING:"),
    recode_values_tidy
  )) |>
  mutate(across(
    contains("AFTER TRAINING:"),
    recode_values_tidy
  )) |>
  mutate(across(
    contains("How comfortable are you delivering serious news"),
    recode_values_tidy
  )) |>
  mutate(across(
    contains("how comfortable were you delivering serious news"),
    recode_values_tidy
  ))
export(complete_data1, "complete_data1.rds")
#summarize data in n(%)
complete_data1 |>
  tbl_summary(digits = list(all_categorical() ~ c(0, 1)))
```

## Bivariate analysis-McNemar's test

dataset: binary data (complete_data1)

### compare pre and post

```{r}
#| echo: false
#McNemar's test
#automation
# Identify pairs of questions
paired_questions <- c("Convey serious news about a patient's illness to the patient or family", 
                      "Convey prognosis to a patient or family member",
                      "Recognize emotion and respond with empathy to a patient or family member",
                      "Lead a family meeting",
                      "Manage conflict that arises during a difficult discussion",
                      "Respond to patients who deny the seriousness of their illness",
                      "Respond to patients who want treatments you believe are not indicated"
                      # ... add other question stems here
)

# Loop through pairs and perform McNemar's tests
results <- vector("list", length = length(paired_questions))
names(results) <- paired_questions

for (question_stem in paired_questions) {
  before_col <- paste("Pre:", question_stem)
  after_col <- paste("AFTER TRAINING:", question_stem)
  
  # Create a contingency table for the variables
  contingency_table <- table(complete_data1[[before_col]], complete_data1[[after_col]])
  print(contingency_table)
  
  # Run McNemar's test only if the contingency table is a 2x2 table
  if(is.matrix(contingency_table) && nrow(contingency_table) == 2 && ncol(contingency_table) == 2){
    mcnemar_result <- mcnemar.test(contingency_table)
    print(mcnemar_result)
    
    # Storing the result if the test is performed
    results[[question_stem]] <- mcnemar_result
  } else {
    print(paste("The contingency table for", question_stem, "is not suitable for McNemar's test."))
    # Store NA to indicate the test was not performed
    results[[question_stem]] <- NA
  }
}

# To see the results of the first test, if it was performed
if (!is.na(results[[1]])) {
  print(results[[1]])
} else {
  print("Result for the first question stem is not available because the test was not performed.")
}
```

### compare pre-retro and post

```{r}
#missing data
complete_data1_nomiss<- complete_data1 |>
  filter(!id %in% c("0612", "4009", "2995") )
library(gtsummary)
complete_data1_nomiss |>
  tbl_summary()
```

```{r}
# Loop through pairs and perform McNemar's tests
results <- vector("list", length = length(paired_questions))
names(results) <- paired_questions

for (question_stem in paired_questions) {
  before_col <- paste("BEFORE TRAINING:", question_stem)
  after_col <- paste("AFTER TRAINING:", question_stem)
  
  # Create a contingency table for the variables
  contingency_table <- table(complete_data1_nomiss[[before_col]], complete_data1_nomiss[[after_col]])
  print(contingency_table)
  
  # Run McNemar's test only if the contingency table is a 2x2 table
  if(is.matrix(contingency_table) && nrow(contingency_table) == 2 && ncol(contingency_table) == 2){
    mcnemar_result <- mcnemar.test(contingency_table)
    print(mcnemar_result)
    
    # Storing the result if the test is performed
    results[[question_stem]] <- mcnemar_result
  } else {
    print(paste("The contingency table for", question_stem, "is not suitable for McNemar's test."))
    # Store NA to indicate the test was not performed
    results[[question_stem]] <- NA
  }
}
# To see the results of the first test, if it was performed
if (!is.na(results[[1]])) {
  print(results[[1]])
} else {
  print("Result for the first question stem is not available because the test was not performed.")
}
```

```{r}

```

### compare pre and pre-retro

```{r}
for (question_stem in paired_questions) {
  before_col <- paste("Pre:", question_stem)
  retro_col <- paste("BEFORE TRAINING:", question_stem)
  
  # Create a contingency table for the variables
  contingency_table <- table(complete_data1_nomiss[[before_col]], complete_data1_nomiss[[retro_col]])
  print(contingency_table)
  
  # Run McNemar's test only if the contingency table is a 2x2 table
  if(is.matrix(contingency_table) && nrow(contingency_table) == 2 && ncol(contingency_table) == 2){
    mcnemar_result <- mcnemar.test(contingency_table)
    print(mcnemar_result)
    
    # Storing the result if the test is performed
    results[[question_stem]] <- mcnemar_result
  } else {
    print(paste("The contingency table for", question_stem, "is not suitable for McNemar's test."))
    # Store NA to indicate the test was not performed
    results[[question_stem]] <- NA
  }
}

```

## Bivariate analysis-Wilcoxon Signed-Rank Test

dataset: likert scale data (complete_data)

### compare pre and post

```{r}
paired_questions <- c(
  "Convey serious news about a patient's illness to the patient or family",
  "Convey prognosis to a patient or family member",
  "Recognize emotion and respond with empathy to a patient or family member",
  "Lead a family meeting",
  "Manage conflict that arises during a difficult discussion",
  "Respond to patients who deny the seriousness of their illness",
  "Respond to patients who want treatments you believe are not indicated"
  # ... add other question stems here
)

# Initialize a list to store the results
wilcoxon_results <- list()

for (question_stem in paired_questions) {
  before_col <- paste("Pre:", question_stem)
  after_col <- paste("AFTER TRAINING:", question_stem)
  
  # Select the scores for before and after training
  before_scores <- complete_data[[before_col]]
  after_scores <- complete_data[[after_col]]
  
  # Calculate median, Q1 and Q3 for before and after scores
  before_median <- median(before_scores, na.rm = TRUE)
  after_median <- median(after_scores, na.rm = TRUE)
  before_quantiles <- quantile(before_scores, probs = c(0.25, 0.75), na.rm = TRUE)
  after_quantiles <- quantile(after_scores, probs = c(0.25, 0.75), na.rm = TRUE)
  
  # Perform the Wilcoxon Signed-Rank Test
  test_result <- wilcox.test(before_scores, after_scores, paired = TRUE, exact = FALSE, correct = TRUE)
  
  # Summarize and store the results
  wilcoxon_results[[question_stem]] <- list(
    before_summary = summary(before_scores),
    after_summary = summary(after_scores),
    before_median_iqr = sprintf("Median (Q1, Q3): %s (%s, %s)", before_median, before_quantiles[1], before_quantiles[2]),
    after_median_iqr = sprintf("Median (Q1, Q3): %s (%s, %s)", after_median, after_quantiles[1], after_quantiles[2]),
    test_result = test_result
  )
# Optional: Print the results for each test
  cat("Wilcoxon Signed-Rank Test for:", question_stem, "\n")
  cat("Before Training Median (Q1, Q3):", wilcoxon_results[[question_stem]]$before_median_iqr, "\n")
  cat("After Training Median (Q1, Q3):", wilcoxon_results[[question_stem]]$after_median_iqr, "\n")
  print(test_result)
  cat("\n")
}
```

### compare pre-retro and post

```{r}
#missing data
complete_data_nomiss<- complete_data |>
  filter(!id %in% c("0612", "4009", "2995") )
library(gtsummary)
complete_data_nomiss |>
  tbl_summary()
```

```{r}
#automation
paired_questions <- c(
  "Convey serious news about a patient's illness to the patient or family",
  "Convey prognosis to a patient or family member",
  "Recognize emotion and respond with empathy to a patient or family member",
  "Lead a family meeting",
  "Manage conflict that arises during a difficult discussion",
  "Respond to patients who deny the seriousness of their illness",
  "Respond to patients who want treatments you believe are not indicated"
  # ... add other question stems here
)

# Initialize a list to store the results
wilcoxon_results <- list()

for (question_stem in paired_questions) {
  before_col <- paste("BEFORE TRAINING:", question_stem)
  after_col <- paste("AFTER TRAINING:", question_stem)
  
  # Select the scores for before and after training
  before_scores <- complete_data_nomiss[[before_col]]
  after_scores <- complete_data_nomiss[[after_col]]
  
  # Calculate median, Q1 and Q3 for before and after scores
  before_median <- median(before_scores, na.rm = TRUE)
  after_median <- median(after_scores, na.rm = TRUE)
  before_quantiles <- quantile(before_scores, probs = c(0.25, 0.75), na.rm = TRUE)
  after_quantiles <- quantile(after_scores, probs = c(0.25, 0.75), na.rm = TRUE)
  
  # Perform the Wilcoxon Signed-Rank Test
  test_result <- wilcox.test(before_scores, after_scores, paired = TRUE, exact = FALSE, correct = TRUE)
  
  # Summarize and store the results
  wilcoxon_results[[question_stem]] <- list(
    before_summary = summary(before_scores),
    after_summary = summary(after_scores),
    before_median_iqr = sprintf("Median (Q1, Q3): %s (%s, %s)", before_median, before_quantiles[1], before_quantiles[2]),
    after_median_iqr = sprintf("Median (Q1, Q3): %s (%s, %s)", after_median, after_quantiles[1], after_quantiles[2]),
    test_result = test_result
  )
    
  # Optional: Print the results for each test
  cat("Wilcoxon Signed-Rank Test for:", question_stem, "\n")
  cat("Before Training Median (Q1, Q3):", wilcoxon_results[[question_stem]]$before_median_iqr, "\n")
  cat("After Training Median (Q1, Q3):", wilcoxon_results[[question_stem]]$after_median_iqr, "\n")
  print(test_result)
  cat("\n")
}
```

### compare pre and pre-retro

```{r}
# add Pre: to pre-workshop questions
complete_data_nomiss <- complete_data_nomiss %>%
  rename(
    `Pre: Convey serious news about a patient's illness to the patient or family` = `Convey serious news about a patient's illness to the patient or family`,
    `Pre: Convey prognosis to a patient or family member` = `Convey prognosis to a patient or family member`,
    `Pre: Recognize emotion and respond with empathy to a patient or family member` = `Recognize emotion and respond with empathy to a patient or family member`,
    `Pre: Lead a family meeting` = `Lead a family meeting`,
    `Pre: Manage conflict that arises during a difficult discussion` = `Manage conflict that arises during a difficult discussion`,
    `Pre: Respond to patients who deny the seriousness of their illness` = `Respond to patients who deny the seriousness of their illness`,
    `Pre: Respond to patients who want treatments you believe are not indicated` = `Respond to patients who want treatments you believe are not indicated`,
    `Pre: Discuss religious or spiritual issues with a patient` = `Discuss religious or spiritual issues with a patient`,
    `Pre: Elicit a patient's personal values and preferences` = `Elicit a patient's personal values and preferences`
  )
# Initialize a list to store the results
wilcoxon_results <- list()

for (question_stem in paired_questions) {
  before_col <- paste("Pre:", question_stem)
  retro_col <- paste("BEFORE TRAINING:", question_stem)
  
  # Select the scores for before and after training
  before_scores <- complete_data_nomiss[[before_col]]
  retro_scores <- complete_data_nomiss[[retro_col]]
  
  # Calculate median, Q1 and Q3 for before and after scores
  before_median <- median(before_scores, na.rm = TRUE)
  retro_median <- median(retro_scores, na.rm = TRUE)
  before_quantiles <- quantile(before_scores, probs = c(0.25, 0.75), na.rm = TRUE)
  retro_quantiles <- quantile(retro_scores, probs = c(0.25, 0.75), na.rm = TRUE)
  
  # Perform the Wilcoxon Signed-Rank Test
  test_result <- wilcox.test(before_scores, retro_scores, paired = TRUE, exact = FALSE, correct = TRUE)
  
  # Summarize and store the results
  wilcoxon_results[[question_stem]] <- list(
    before_summary = summary(before_scores),
    retro_summary = summary(retro_scores),
    before_median_iqr = sprintf("Median (Q1, Q3): %s (%s, %s)", before_median, before_quantiles[1], before_quantiles[2]),
    retro_median_iqr = sprintf("Median (Q1, Q3): %s (%s, %s)", retro_median, retro_quantiles[1], retro_quantiles[2]),
    test_result = test_result
  )
  
  # Optional: Print the results for each test
  cat("Wilcoxon Signed-Rank Test for:", question_stem, "\n")
  cat("Before Training Median (Q1, Q3):", wilcoxon_results[[question_stem]]$before_median_iqr, "\n")
  cat("After Training Retro Median (Q1, Q3):", wilcoxon_results[[question_stem]]$retro_median_iqr, "\n")
  print(test_result)
  cat("\n")
}
```

## Comfortable level

### compare pre and post

```{r}
################comfortable question####################
before_col <- "How comfortable are you delivering serious news to a patient?"
after_col <- "AFTER this training, how comfortable are you delivering serious news to a patient?"

# Create a contingency table for the variables
contingency_table <- table(complete_data1[[before_col]], complete_data1[[after_col]])
print(contingency_table)
# Run McNemar's test
mcnemar_result <- mcnemar.test(contingency_table)
print(mcnemar_result)

# Wilcoxon Signed-Rank Test
before_col <- "How comfortable are you delivering serious news to a patient?" 
after_col <- "AFTER this training, how comfortable are you delivering serious news to a patient?"
before_scores <- complete_data[[before_col]]
after_scores <- complete_data[[after_col]]
# Summary of 'before training' scores
summary(complete_data[[before_col]])
# Summary of 'after training' scores
summary(complete_data[[after_col]])
# Wilcoxon Signed-Rank Test
wilcoxon_result <- wilcox.test(before_scores, after_scores, 
                               paired = TRUE, # paired sample: before and after
                               exact = FALSE, # ties and zero differences exist
                               correct = TRUE) # Apply continuity correction that can be useful for discrete data like Likert scales.
print(wilcoxon_result)
```

### compare pre-retro and post

```{r}
################comfortable question####################
before_col <- "BEFORE this training, how comfortable were you delivering serious news to a patient?"
after_col <- "AFTER this training, how comfortable are you delivering serious news to a patient?"

# Create a contingency table for the variables
contingency_table <- table(complete_data1_nomiss[[before_col]], complete_data1_nomiss[[after_col]])
print(contingency_table)
# Run McNemar's test
mcnemar_result <- mcnemar.test(contingency_table)
print(mcnemar_result)

# Wilcoxon Signed-Rank Test
before_col <- "BEFORE this training, how comfortable were you delivering serious news to a patient?" 
after_col <- "AFTER this training, how comfortable are you delivering serious news to a patient?"
before_scores <- complete_data_nomiss[[before_col]]
after_scores <- complete_data_nomiss[[after_col]]
# Summary of 'before training' scores
summary(complete_data_nomiss[[before_col]])
# Summary of 'after training' scores
summary(complete_data_nomiss[[after_col]])
# Wilcoxon Signed-Rank Test
wilcoxon_result <- wilcox.test(before_scores, after_scores, 
                               paired = TRUE, # paired sample: before and after
                               exact = FALSE, # ties and zero differences exist
                               correct = TRUE) # Apply continuity correction that can be useful for discrete data like Likert scales.
print(wilcoxon_result)
```

## Conclusion

## References
