---
title: "Summary"
author: "Yang Liu"
date: 8/29/2025
format: html
execute:
  echo: false
  warning: false
editor: visual
---

## Title

Clinical Utility and Cost-effectiveness of the Pre-operative Type and Screen in Gynecologic Oncology

## Team

Sarosi, Janke, Brackmann, Morgan

## Background

A **preoperative type and screen** is a common laboratory test ordered before many surgeries to prepare for the possibility that a patient might need a blood transfusion during or after their procedure.

-   **Type**: This refers to determining your blood type (ABO group and Rh factor, e.g., A+, O-, etc.).

-   **Screen**: This checks your blood for unexpected antibodies that could react with donor blood.

Given the risk of intraoperative blood loss during surgery, preoperative type and screen is routinely ordered in GYO patients.

Only a small portion of GYP patients require transfusion of blood products.

Studies have demonstrated that a well-planned use of the pre-operative type and screen in benign gynecologic laparoscopic surgery is both safe and cost-effective. (less type and screen)

**Difference from “Type and Crossmatch”**

-   **Type and Screen:** Blood is *not* specifically reserved for you, just tested in advance. Used when transfusion is possible but not certain.

-   **Type and Crossmatch:** Blood is actually set aside and matched in detail. Used when transfusion is almost certain or urgent.

## Study population

Patients aged \>/= 18 undergoing vulvar, vaginal or laparoscopic Gynecologic Oncology surgery at the University of Michigan. The pre-intervention cohort will range from 1/1/2021 to 12/31/2022. The post-intervention cohort will range from 1/1/2023 to 12/31/2024

```{r}
#| echo: false
library(DiagrammeR)
grViz("digraph flowchart {
      # node definitions with substituted label text
      node [fontname = Helvetica, shape = rectangle]        
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']

      # edge definitions with the node IDs
      tab1 -> tab2 -> tab3 -> tab4 -> tab5;
      }

      [1]: 'GYO patients 1/1/2021-12/31/2024 (N=2367)'
      [2]: 'Intervention December 2022'
      [3]: 'Exclusion.....'
      [4]: 'Pre-intervention cohort 1/1/2021-12/31/2022 N='
      [5]: 'Post-intervention cohort 1/1/2023-12/31/2024 N='
      ")
```

## Objective

Through a pre-post study design, we aim to characterize the utility of the preoperative type and screen (T&S) in patients undergoing Gynecologic Oncology surgery.

## Prediction

1.  Intervention lowers the rate of T&S
2.  Patients who had open surgery are more likely to have pre-operative T&S.

## Primary outcome

Pre-post changes in T&S ordering practices for vulvar, vaginal and laparoscopic Gynecologic Oncology surgeries

## Secondary outcome

Pre-post changes in morbidity, including rates of conversion from laparoscopic to laparotomy, estimated blood loss, rates of conversion from T&S to type and cross (T&C) without a transfusion, rates of perioperative (intra-operatively or post-operatively during the same hospitalization) transfusions, rates of urgent transfusions of un-crossmatched blood, and mortality rates.

Cost savings associated with this intervention.

## Variables of interest

-Demographic characteristics: age, race, BMI, insurance

-Clinical characteristics (comorbidity): anemia, hypertension, hyperlipidemia, diabetes, hyperthyroidism, hypothyroidism, COPD, asthma, OSA, GERD, diverticulosis, alcohol use, tobacco use, prior exposure to chemotherapy

-Surgical characteristics: surgical type, surgical indication, length of stay after surgery (days), laparoscopy to laparotomy conversion, EBL, conversion to T&C without transfusion, perioperative transfusion, un-crossmatched transfusion, mortality

-T&S characteristics: pre-operative T&S without transfusion, intra-operative T&S without transfusion, post-operative T&S without transfusion, T&C without transfusion

## Raw Data Description

Data source: Michart

Data collection: by Ken

1.  Available: sheet "QI", variable "indicatorWlist": ICU admission, transfusion perioperative, transfusion not perioperative. all listed Michart_Log_Id had transfusion if the values are "transfusion perioperative" or "transfusion not perioperative". not listed had no transfusion. ICU admission had no transfusion? "transfusion not periopetative" means UNKNOWN pre-operative transfusion or post-operative? casedate is admission date, not helpful in defining type of transfusion.

    sheet "Orders", all listed Michart_Log_Id had T&S orders. what is "NO TS REQUIRED FOR OR" (patients don't need T&S-intervention?), variable "T&S_On_Surgery_Date" is intra-operative T&S, variable "30days_Prior" is pre-operative T&S, variable "30days_Post" is post-operative T&S.

## Data management(Import, Cleaning, Variable Creation)

### Load Packages

```{r}
library(readxl)
library(tidyverse) #for import, cleaning, wrangling
library(gtsummary) #for summarizing
library(labelled) #for variable labelling
library(rio) #import/export 
library(janitor) #for cleaning and tabulation helpers
library(visdat) #data visualization for missingness
library(hmisc) #for variable labelling
```

### Import data

```{r}
#--- Import raw datasets from Excel ---
#raw data structure: 
#sheet "Cohort" demographics
#sheet "Categories" surgical type
#sheet "Dx" surgical indication, comorbidities
#sheet "Orders" Type and screen orders
#sheet "T&S Done" Type and screen done
#sheet "QI" QI indicators (death, surgery conversion, transfusion, blood loss, etc)

cohort <- read_excel("rawdata/Type & Screen Data 022626.xlsx", 
    sheet = "Cohort") #n=2328
category <- read_excel("rawdata/Type & Screen Data 022626.xlsx", 
    sheet = "T&S + Comorbidities") #n=877
dx <- read_excel("rawdata/Type & Screen Data 022626.xlsx", 
    sheet = "Dx") #n=10134
qi <- read_excel("rawdata/Type & Screen Data 022626.xlsx", 
    sheet = "QI") #n=1155
```

### Initial Data Check and Cleaning

```{r}
#--- Basic structure and missingness checks ---
str(data)          # Data structure
head(data)         # First 6 rows
dim(data)          # Number of rows & columns
names(data)        # Variable names
visdat::vis_miss(data) # Visualize missing data

#replace empty string with NA, whole dataset, all columns
data[data == ''] <- NA 
#replace only on selected columns
data <- data |> mutate_at('selected var', ~na_if(., '')) 

#--- basic cleaning ---
cohort <- cohort|> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))
category <- category|> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))
dx <- dx |> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))
qi <- qi |> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))

#--- Custom NA summary function ---
check_na <- function(df) {
  total_na <- sum(is.na(df))
  na_per_column <- colSums(is.na(df))
  list(total_na = total_na, na_per_column = na_per_column)
}
check_na(case_complete)

#--- check MRN multiple entries ---
cohort |>
  get_dupes(mrn)    #mrn multiple entries
#check case_id multiple entries
cohort |>
  get_dupes(case_id)   #case_id unique
#check michart_log_id multiple entries
cohort |>
  get_dupes(michart_log_id)  #michart_log_id unique
```

### Merge Demographics and Comorbidities

```{r}
#organize comorbidities
comor_codes <- list(
  obesity           = c("E66.9", "E66.813", "E66.1", "E66.09", "E66.01"),
  anemia            = c("D64.9", "D50.0", "D64.81"),
  htn               = c("I10"),
  hld               = c("E78.5", "E78.49", "E78.2", "E78.00"),
  diabetes          = c("E11.9", "E11.8", "E11.69", "E11.65", "E11.59", "E11.49", "E11.43", "E11.42", "E11.40", "E11.3299", "E11.319", "E11.29", "E11.22", "E11.21", "E10.9", "E10.8", "E10.69", "E10.65", "E10.3293", "E10.22"),
  hypothyroidism    = c("E89.0", "E03.9", "E03.8"),
  copd              = c("J44.9", "J44.1", "J43.9", "J42"),
  asthma            = c("J45.990", "J45.909", "J45.50", "J45.40", "J45.30", "J45.20"),
  osa               = c("G47.33", "G47.34", "G47.30"),
  gerd              = c("K21.9", "K21.00"),
  ckd               = c("N18.9", "N18.6", "N18.5", "N18.4", "N18.32", "N18.31", "N18.30", "N18.2", "N18.1", "I12.9", "I13.2", "I13.0", "E11.22"),
  acutekidneyfailure= c("N17.9", "N17.0"),
  heartfailure      = c("I50.9", "I500.42", "I50.33", "I50.32", "I50.30", "I50.22", "I50.20"),
  afib              = c("I48.92", "I48.91", "I48.21", "I48.20", "I48.19", "I48.11", "I48.0"),
  endometriosis     = c("N80.9", "N80.8", "N80.512", "N80.3A3", "N80.341", "N80.321", "N80.3", "N80.202", "N80.2", "N80.129", "N80.123", "N80.122", "N80.121", "N80.111", "N80.103", "N80.102", "N80.101", "N80.1", "N80.0"),
  anticoagulant     = c("Z79.01"),
  tobaccouse        = c("Z72.0", "Z87.891", "F17.210", "F17.200")
)
# Unnest and tidy logic:
dx_flags <- map_dfc(names(comor_codes), function(comorb) {
  dx %>%
    mutate(flag = as.integer(dx_cd %in% comor_codes[[comorb]])) %>%
    group_by(michart_log_id) %>%
    summarise(!!comorb := max(flag)) %>%
    ungroup() %>%
    select(-michart_log_id)
})

# Reattach michart_log_id
output <- dx %>%
  distinct(michart_log_id) %>%
  bind_cols(dx_flags)
# spot check using any comorbidity
output %>%
  filter(htn==1)|>
  select(michart_log_id, htn)|>
  print(n=100)
#merge cohort and dx
output$michart_log_id <- as.character(output$michart_log_id)
case_complete <- left_join(cohort, output, by = "michart_log_id") #n=2328 unique case
```

### Merge Diagnosis

```{r}
# --- Organize diagnosis ---
pathology_codes <- list(
  vulvar_benign = c("N90.89", "N90.7", "N90.69", "N90.60", "N76.6", "N76.2", "N75.1", "N75.0", "D28.0"),
  vulvar_precancer = c("N90.4", "N90.3", "N90.1", "N90.0", "L90.0"),
  vulvar_cancer = c("D07.1", "C51.9", "C51.8", "C51.2", "C51.1", "C51.0"),
  vaginal_benign = c("N89.8", "N84.2", "N76.5", "D28.1"),
  vaginal_precancer = c("N89.3", "N89.1", "N89.0"),
  vaginal_cancer = c("D07.2", "C52"),
  cervical_benign = c("N88.8", "N84.1", "N72", "D26.0"),
  cervical_precancer = c("N87.9", "N87.1", "N87.0", "R87.623", "R87.622", "R87.621", "R81.613", "R87.611", "R87.610"),
  cervical_cancer = c("Z85.41", "D06.9", "D06.0", "C53.9", "C53.8", "C53.1", "C53.0"),
  uterine_benign = c("N85.9", "N85.2", "N85.01", "N84.0", "D26.9", "D26.7", "D26.1", "D25.9", "D25.2", "D25.1", "D25.0", "D39.0"),
  uterine_precancer = c("N85.00", "N85.02"),
  uterine_cancer = c("Z85.42", "D07.00", "C55", "C54.9", "C54.8", "C54.2", "C54.1", "C54.0"),
  ovarianft_peritoneal_benign = c("D20.1", "D20.0", "D19.1", "N83.9", "N83.8", "N83.519", "N83.512", "N83.511", "N83.292", "N83.209", "N83.202", "N83.201", "N83.12", "N83.11", "N83.02", "N83.01", "N83.00", "N70.93", "N70.92", "N70.91", "N70.13", "N70.11", "N70.01", "D39.12", "D39.11", "D39.10", "D27.9", "D27.1", "D27.0"),
  ovarianft_peritoneal_cancer = c("C48.1", "Z85.43", "C79.63", "C57.4", "C57.02", "C57.01", "C57.00", "C56.9", "C56.3", "C56.2", "C56.1"),
  gi_benign = c("D13.39", "D13.1", "D12.9", "D12.8", "D12.7", "D12.6", "D12.5", "D12.3", "D12.1", "D12.0", "D3A.020", "D37.3"),
  gi_cancer = c("D01.3", "C21.0", "C20", "C19", "C18.9", "C18.7", "C18.4", "C18.2", "C18.1", "C17.2"),
  femalegenital_unspecified_benign = c("D49.59", "D28.9", "D28.7", "D49.9", "D48.9", "D48.7", "D36.9", "D36.7", "D28.2"),
  femalegenital_unspecified_cancer = c("Z85.40", "Z85.44", "D07.39", "C80.1", "C79.9", "C79.89", "C79.82", "C57.9"),
  placental_benign = c("D39.2", "O43.219", "O43.213", "O43.212", "O43"),
  placental_cancer = c("C58"),
  lymphnode_benign = c("D36.0"),
  urinarytract_benign = c("D30.4", "D30.3"),
  urinarytract_cancer = c("D09.19", "C68.0", "C67.8", "C67.5", "C67.3"),
  melanoma = c("D03.8", "D03.59", "C43.59", "C43.9"),
  skin_cancer = c("C44.99", "C44.599", "C44.520", "C44.519")
)
# Unnest and tidy logic:
path_flags <- map_dfc(names(pathology_codes), function(path) {
  dx %>%
    mutate(flag = as.integer(dx_cd %in% pathology_codes[[path]])) %>%
    group_by(michart_log_id) %>%
    summarise(!!path := max(flag)) %>%
    ungroup() %>%
    select(-michart_log_id)
})

# Reattach michart_log_id
output_path <- dx %>%
  distinct(michart_log_id) %>%
  bind_cols(path_flags)
# spot check 
output_path %>%
  filter(vulvar_cancer==1)|>
  select(michart_log_id, vulvar_cancer)|>
  print(n=100)
#--- Organize cancer/benign ---
cancer_benign_codes <- list(
  cancer = c(
    "D07.1", "C51.9", "C51.8", "C51.2", "C51.1", "C51.0", 
    "D07.2", "C52", "Z85.41", "D06.9", "D06.0", "C53.9", 
    "C53.8", "C53.1", "C53.0", "Z85.42", "D07.00", "C55", 
    "C54.9", "C54.8", "C54.2", "C54.1", "C54.0", "C48.1", 
    "Z85.43", "C79.63", "C57.4", "C57.02", "C57.01", "C57.00", 
    "C56.9", "C56.3", "C56.2", "C56.1", "D01.3", "C21.0", 
    "C20", "C19", "C18.9", "C18.7", "C18.4", "C18.2", 
    "C18.1", "C17.2", "Z85.40", "Z85.44", "D07.39", "C80.1", 
    "C79.9", "C79.89", "C79.82", "C57.9", "C58", "D09.19", 
    "C68.0", "C67.8", "C67.5", "C67.3", "D03.8", "D03.59", 
    "C43.59", "C43.9", "C44.99", "C44.599", "C44.520", "C44.519"
  ),
  benign = c(
    "N90.89", "N90.7", "N90.69", "N90.60", "N76.6", "N76.2", 
    "N75.1", "N75.0", "D28.0", "N90.4", "N90.3", "N90.1", 
    "N90.0", "L90.0", "N89.8", "N89.3", "N89.1", "N89.0", 
    "N88.8", "N84.1", "N72", "D26.0", "N87.9", "N87.1", 
    "N87.0", "R87.623", "R87.622", "R87.621", "R81.613", 
    "R87.611", "R87.610", "N85.9", "N85.2", "N85.01", "N84.0", 
    "D26.9", "D26.7", "D26.1", "D25.9", "D25.2", "D25.1", 
    "D25.0", "D39.0", "N85.00", "N85.02", "D20.1", "D20.0", 
    "D19.1", "N83.9", "N83.8", "N83.519", "N83.512", "N83.511", 
    "N83.292", "N83.209", "N83.202", "N83.201", "N83.12", "N83.11", 
    "N83.02", "N83.01", "N83.00", "N70.93", "N70.92", "N70.91", 
    "N70.13", "N70.11", "N70.01", "D39.12", "D39.11", "D39.10", 
    "D27.9", "D27.1", "D27.0", "D13.39", "D13.1", "D12.9", 
    "D12.8", "D12.7", "D12.6", "D12.5", "D12.3", "D12.1", 
    "D12.0", "D3A.020", "D37.3", "D49.59", "D28.9", "D28.7", 
    "D49.9", "D48.9", "D48.7", "D36.9", "D36.7", "D28.2", 
    "D39.2", "O43.219", "O43.213", "O43.212", "O43", "D36.0", 
    "D30.4", "D30.3"
  )
)
# Unnest and tidy logic:
cancer_flags <- map_dfc(names(cancer_benign_codes), function(cancer) {
  dx %>%
    mutate(flag = as.integer(dx_cd %in% cancer_benign_codes[[cancer]])) %>%
    group_by(michart_log_id) %>%
    summarise(!!cancer := max(flag)) %>%
    ungroup() %>%
    select(-michart_log_id)
})

# Reattach michart_log_id
output_cancer <- dx %>%
  distinct(michart_log_id) %>%
  bind_cols(cancer_flags)
#--- Merge case, output_path, output_cancer by michart_log_id
output_path$michart_log_id <- as.character(output_path$michart_log_id)
output_cancer$michart_log_id <- as.character(output_cancer$michart_log_id)
case_complete <- left_join(case_complete, output_path, by = "michart_log_id") #n=2328 unique case 
case_complete <- left_join(case_complete, output_cancer, by = "michart_log_id") #n=2328 unique case for outcome
```

### Merge QI data

```{r}
### organize QI data
#create a list
indicator_codebook <- list(
  conversion = "GYN: 14. Conversion from laparoscopy to laparotomy",
  death = "GYN:1:Death",
  transfusion_periop = "GYN:12.1:Bleeding: transfusion, perioperative",
  ebl_1000 = "GYN:12.41:Bleeding: BL >= 1,000cc",
  ebl_1500 = "GYN:12.42:Bleeding: BL >= 1,500cc",
  ebl_2000 = "GYN:12.43:Bleeding: BL >= 2,000cc",
  ebl_2500 = "GYN:12.44:Bleeding: BL >= 2,500cc",
  injury_major = "GYN:5.3:Surgical injury: major vessel",
  transfusion_periop_pph = "OB:12.1:Bleeding: transfusion, perioperative OR PPH",
  transfusion_periop_4 = "OB:12.6:Bleeding: transfusion, perioperative or PPH, 4 or more units packed RBC"
)
#create indicator columns
wanted_vars <- names(indicator_codebook)
qi_flagged <- map_dfc(
  wanted_vars,
  ~ qi %>%
      mutate(flag = as.integer(indicator_wlist == indicator_codebook[[.x]])) %>%
      group_by(michart_log_id) %>%
      summarise(!!.x := max(flag), .groups = "drop") %>%
      select(-michart_log_id)
)

result <- qi %>% distinct(michart_log_id) %>% bind_cols(qi_flagged)
print(result) #n=471 obs
# merge qi data
result$michart_log_id <- as.character(result$michart_log_id)

case_complete <- left_join(case_complete, result, by = "michart_log_id") #n=2328 cases
```

### Merge Surgical type and T&S data

```{r}
#merge category} 
category <- category |>   
  filter(!is.na(preop_t_s))  
category$michart_log_id <- as.character(category$michart_log_id)  
case_mis <- left_join(category, case_complete,by = "michart_log_id") #n=877 unique case

```

### **Variable Transformation**

```{r}
####################tidy variables################################
#calculate bmi
case_mis <- case_mis |>
  mutate(height_m=height_in*0.0254,
         bmi=weight_kg/height_m ^ 2,
         bmigroup = case_when(bmi < 18.5 ~ "Under weight",
                              bmi < 25 ~ "Normal weight",
                              bmi < 30 ~ "Overweight",
                              bmi >= 30 ~ "Obese"))
case_mis$bmigroup<-  forcats::fct_relevel(case_mis$bmigroup, "Under weight", "Normal weight", "Overweight","Obese")
# create variable 
#relevel
case_mis$cohort <- forcats::fct_relevel(case_mis$cohort, "Pre-Intervention", "Post-Intervention")
# create binary variable 
case_mis <- case_mis |>
  mutate(ts = if_else(preop_t_s == "N", 0, 1))
label(case_mis$ts)<- "Type & Screen"
# convert transfusion_periop NA to 0
case_mis <- case_mis |>
  mutate(transfusion_periop=as.numeric(transfusion_periop))|>
  mutate(transfusion_periop=case_when(is.na(transfusion_periop) ~ 0,
                       TRUE ~ transfusion_periop))
label(case_mis$transfusion_periop)<- "Periop transfusion"

```

### Fix missing data

```{r}
#fix missing diagnosis
case_mis <- case_mis |>
  mutate(cancertype=case_when(diagnosis_2=="cancer" ~ "Cancer",
                              TRUE ~ diagnosis_2))
#collect mrns that have missing ebl
case_noebl <- case_mis|>
  filter(is.na(ebl))|>
  select(mrn.x, michart_log_id)
export(case_noebl, "rawdata/case_noebl.csv")
#import ebl data
case_ebl <- read_csv("rawdata/case_noebl.csv")
#old codes: merge
case_ebl$michart_log_id <- as.character(case_ebl$michart_log_id) 
case_mis <- case_mis %>%
  left_join(case_ebl, by = "michart_log_id", suffix = c("_1", "_2")) %>%
  mutate(ebl = coalesce(ebl_1, ebl_2)) 
#updated
case_mis$ebl.y <- as.character(case_mis$ebl.y) 
case_mis <- case_mis %>%
  mutate(ebl = coalesce(ebl.x, ebl.y)) 
ebl <- case_mis |>
  select(ebl.x, ebl.y,ebl)
case_mis <- case_mis |>
  mutate(
    ebl = if_else(ebl == "N/A", "0", ebl),  # Everything is a string now!
    ebl = as.numeric(ebl)                   # Convert to numeric
  )
```

### **Variable Labeling**

```{r}
label(case_mis$age.y)<- "Age"
label(case_mis$race.y)<- "Race"
label(case_mis$ethnicity.y)<- "Ethnicity"
label(case_mis$bmi)<- "BMI"
label(case_mis$ebl)<- "EBL"
```

### Create Dataset for Demographic Table

```{r}
#############create dataset for demographic############################
#keep single MRN entry by first visit
mrn_mis <- case_mis |>
  arrange(mrn.x, casedate) |>
  distinct(mrn.x, .keep_all = TRUE)   #n=871

```

### **Save Final Data**

```{r}
#save
export(case_complete, "case_complete.rds") #unique case n=2328
export(case_mis, "case_mis.rds") #unique case n=877
export(mrn_mis, "mrn_mis.rds") #unique mrn n=871
```

## Descriptive summary

Table 1 Demographics (by unique MRN, first visit)

```{r}
# descriptive summary
mrn_mis |>  
  select(age.y, race.y, ethnicity.y, bmi, ebl, alcohol_use:tobaccouse, cohort) |>
  tbl_summary(
    #missing = "no",
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_overall()
```

Table 2 Type and Screen rates (by case)

All cases

```{r}
#Table 2 Type and Screen rates (by case)
case_mis |>  
  select(cohort, ts, transfusion_periop) |>
  tbl_summary(
    #missing = "no",
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()
```

by diagnosis

```{r}
#by diagnosis
case_mis |> 
  filter(cancertype=="Benign/unknown")|>
  select(cohort, ts) |>
  tbl_summary(
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()

case_mis |> 
  filter(cancertype=="Cancer")|>
  select(cohort, ts) |>
  tbl_summary(
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()

case_mis |> 
  filter(cancertype=="Pre-cancer")|>
  select(cohort, ts) |>
  tbl_summary(
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()
```

Data discrepancy

## Bivariate analysis

```{r}
#| echo: false
data2 |>
  select( var of interest) |>
  tbl_summary(
    by=var,
    # missing = "no",
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()|>
modify_header(
    stat_1 ~ "**something**",
    stat_2 ~ "**something**"
  ) |> 
  modify_spanning_header(all_stat_cols() ~ "**whatever**")|>
  modify_caption("**Table 2. whatever Characteristics by whatever**")
```

## Conclusion

## References

Data Discrepancy: raw data sheet "Cohort" demographics and sheet "Categories" surgical type, some don't match: total 271 cases are not listed in categories.

```{r}
case_notype <- category|>
  filter(is.na(column1))|>
  select(mrn, michart_log_id)
export(case_notype, "case_notype.csv")




```

## Abstract

SGO 2026 Winter Meeting (not accepted)

|  |  |
|------------------------------------|------------------------------------|
| **Title:** | Reducing Unnecessary Preoperative Type and Screen Utilization: A Quality Improvement Initiative in Minimally Invasive Gynecologic Oncology Surgery |
| **Objectives:** | To evaluate the impact of reducing unnecessary preoperative type and screen (T&S) testing for patients undergoing minimally invasive gynecologic oncology surgery. |
| **Methods:** | We conducted a pre/post quality improvement study at a single academic medical center to assess outcomes of implementing selective use of the preoperative T&S. In December 2022, gynecologic oncology surgeons were presented with institutional data demonstrating that transfusion rates for minimally invasive surgeries (MIS) were 1.8% and were subsequently encouraged to discontinue routine use of the preoperative T&S for these cases. The pre-intervention period was January 2021-December 2022 and post-intervention period was January 2023-December 2024. A random selection of MIS cases (laparoscopic and robotic cases and vaginal hysterectomies) during the study period were reviewed to compare T&S utilization rates before and after the intervention. T&S utilization was stratified by preoperative diagnosis (cancer, pre-cancer, or benign/unknown). Perioperative rates of transfusion and transfusion reactions were assessed. Cost savings of omitting preoperative T&S were calculated. |
| **Results:** | Of the 381 MIS cases identified during the study period, 188 occurred before and 193 after the intervention. Preoperative T&S use significantly decreased post-intervention from 96.3% (181/188) to 42.5% (82/193) (p \< 0.001). This reduction was consistent across all preoperative diagnosis types: cancer (96.1% pre-intervention to 45.0% post-intervention), pre-cancer (94.0% pre-intervention to 40.4% post-intervention), and benign/unknown diagnoses (100.0% pre-intervention to 39.0% post-intervention) (all p\<0.001, Table 1). Perioperative transfusion rates were low in both groups (1.6% pre-intervention vs 2.1% post-intervention, significance not calculated due to low event rates), and no transfusion reactions occurred. From a cost-savings standpoint, avoidance of preoperative T&S in 111 cases during the post-intervention period resulted in an estimated institutional savings of \$27,528, based on a per-test cost of \$248. |
| **Conclusions:** | This quality improvement initiative suggests that selective use of preoperative T&S in minimally invasive gynecologic oncology surgery is feasible, safe and cost-effective. |

## Data analysis round 2

```{r}
#import raw data
category <- read_excel("rawdata/Type & Screen Data_121725.xlsx", 
    sheet = "Categories - T&S ") #n=878
category <- category|> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))
#merge category
category <- category |>   
  filter(!is.na(preop_t_s))  
category$michart_log_id <- as.character(category$michart_log_id)  
case_mis <- left_join(category, case_complete,by = "michart_log_id") #n=877 unique case
#calculate bmi
case_mis <- case_mis |>
  mutate(height_m=height_in*0.0254,
         bmi=weight_kg/height_m ^ 2,
         bmigroup = case_when(bmi < 18.5 ~ "Under weight",
                              bmi < 25 ~ "Normal weight",
                              bmi < 30 ~ "Overweight",
                              bmi >= 30 ~ "Obese"))
case_mis$bmigroup<-  forcats::fct_relevel(case_mis$bmigroup, "Under weight", "Normal weight", "Overweight","Obese")
# create binary variable 
case_mis <- case_mis |>
  mutate(ts = if_else(preop_t_s == "N", 0, 1))
label(case_mis$ts)<- "Type & Screen"
# convert transfusion_periop NA to 0
case_mis <- case_mis |>
  mutate(transfusion_periop=as.numeric(transfusion_periop))|>
  mutate(transfusion_periop=case_when(is.na(transfusion_periop) ~ 0,
                       TRUE ~ transfusion_periop))
label(case_mis$transfusion_periop)<- "Periop transfusion"
#relevel
case_mis$cohort <- forcats::fct_relevel(case_mis$cohort, "Pre-Intervention", "Post-Intervention")
#fix missing diagnosis
case_mis <- case_mis |>
  mutate(cancertype=case_when(diagnosis_2=="ba" ~ "Benign/unknown",
                              diagnosis_2=="ca" ~ "Cancer",
                              TRUE ~ diagnosis_2))
#keep single MRN entry by first visit
mrn_mis <- case_mis |>
  arrange(mrn.x, casedate) |>
  distinct(mrn.x, .keep_all = TRUE)   #n=871
export(case_mis, "case_mis.rds") #unique case n=871
export(mrn_mis, "mrn_mis.rds") #unique mrn n=877
```

```{r}
# Table 1 Demographics (by unique MRN, first visit)
mrn_mis |>  
  select(age, race, ethnicity, bmi, ebl, alcohol_use:tobaccouse, cohort) |>
  tbl_summary(
    #missing = "no",
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_overall()
#Table 2 Type and Screen rates (by case)
case_mis |>  
  select(cohort, ts, transfusion_periop) |>
  tbl_summary(
    #missing = "no",
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()
#by diagnosis
case_mis |> 
  filter(cancertype=="Benign/unknown")|>
  select(cohort, ts) |>
  tbl_summary(
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()

case_mis |> 
  filter(cancertype=="Cancer")|>
  select(cohort, ts) |>
  tbl_summary(
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()

case_mis |> 
  filter(cancertype=="Pre-cancer")|>
  select(cohort, ts) |>
  tbl_summary(
    by=cohort,
    digits = list(all_categorical() ~ c(0, 1)))|>
  add_p()
```

notes: ebl data no updates

### 2/25/2026

877 vs 878 cases-missing one case

```{r}
#identify missing case and missing data
category1 <- read_excel("rawdata/Type & Screen Data 022526.xlsx", 
    sheet = "T&S + Comorbidities") #n=878
category1 <- category1|> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))
category1$michart_log_id <- as.character(category1$michart_log_id)
#category n=877
#case_complete n=2328
missingcase <- anti_join(category1, category, by="michart_log_id")
#michart id: 3435312
#missing data
case_no <- category1|>
  filter(is.na(diagnosis_2))|>
  select(mrn, michart_log_id)
#mrn, michart_log_id
1
014225766
3435312
2
101561174
3087333
```

Analysis

```{r}
#import raw data
category <- read_excel("rawdata/Type & Screen Data 022626.xlsx", 
    sheet = "T&S + Comorbidities") #n=877
category <- category|> 
  janitor::clean_names() |> 
  janitor::remove_empty(which = c("rows", "cols"))
#merge category
category <- category |>   
  filter(!is.na(preop_t_s))  
category$michart_log_id <- as.character(category$michart_log_id)  
case_mis <- left_join(category, case_complete,by = "michart_log_id") #n=877 unique case
#calculate bmi
case_mis <- case_mis |>
  mutate(height_m=height_in*0.0254,
         bmi=weight_kg/height_m ^ 2,
         bmigroup = case_when(bmi < 18.5 ~ "Under weight",
                              bmi < 25 ~ "Normal weight",
                              bmi < 30 ~ "Overweight",
                              bmi >= 30 ~ "Obese"))
case_mis$bmigroup<-  forcats::fct_relevel(case_mis$bmigroup, "Under weight", "Normal weight", "Overweight","Obese")
# create binary variable 
case_mis <- case_mis |>
  mutate(ts = if_else(preop_t_s == "N", 0, 1))
label(case_mis$ts)<- "Type & Screen"
# convert transfusion_periop NA to 0
case_mis <- case_mis |>
  mutate(transfusion_periop=as.numeric(transfusion_periop))|>
  mutate(transfusion_periop=case_when(is.na(transfusion_periop) ~ 0,
                       TRUE ~ transfusion_periop))
label(case_mis$transfusion_periop)<- "Periop transfusion"
#relevel
case_mis$cohort <- forcats::fct_relevel(case_mis$cohort, "Pre-Intervention", "Post-Intervention")
#fix missing diagnosis
case_mis <- case_mis |>
  mutate(cancertype=case_when(diagnosis_2=="ba" ~ "Benign/unknown",
                              diagnosis_2=="ca" ~ "Cancer",
                              TRUE ~ diagnosis_2))
#keep single MRN entry by first visit
mrn_mis <- case_mis |>
  arrange(mrn.x, casedate) |>
  distinct(mrn.x, .keep_all = TRUE)   #n=871
export(case_mis, "case_mis.rds") #unique case n=871
export(mrn_mis, "mrn_mis.rds") #unique mrn n=877
```

Manuscript
